---
title: "Event subsetting"
output: html_notebook
---

This file contains functions and a main script that can be used to:
  1. Extract GPS track segments associated with specific behaviors using HMM states.
  2. Identify GPS track segments with movement patterns indicative of "boat events" (e.g.,             back-and-forth movement).
  3. Compare detected boat events to known events to assess detection recall.
  4. Visualize selected GPS track segments for precision evaluation.

The input CSV file should have the following columns:
1. time (POSIXct): Timestamp of each data point in the format "YYYY-MM-DD HH:MM:SS".
2. States (string): Behavior state associated with the data point (e.g., "Trawler").
3. x (numeric): UTM X-coordinate of the data point.
4. y (numeric): UTM Y-coordinate of the data point.

An additional file with pre-selected events is necessary for the recall evaluation along with the event_summary_df which can be created using event_analysis.Rmd

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(lubridate)
```

# extract GPS track segments with specific HMM states labels
```{r}
filter_states <- function(gps_data) {

  # Filter the dataset to include only boat points (state 6)
  boat_data <- gps_data %>% filter(States %in% c("Trawler", "Shrimp boat", "Exploratory Flight"))

  # Sort the data by time
  boat_data <- boat_data %>% arrange(time)
  boat_data <- boat_data[!is.na(boat_data$time), ]
  
  # Initialize the event ID
  boat_data$event_id <- 1

  # Loop through the boat_data to assign event IDs
  for (i in 2:nrow(boat_data)) {
    
    # Calculate the time difference between consecutive points
    time_diff <- difftime(boat_data$time[i], boat_data$time[i - 1], units = "mins")
    
    # If the time difference is greater than 30 minutes, increment the event ID
    if (time_diff > 30) {
      boat_data$event_id[i] <- boat_data$event_id[i - 1] + 1
    } else {
      boat_data$event_id[i] <- boat_data$event_id[i - 1]
    }
  }
  
  # Extract the event IDs with their corresponding time ranges
  event_ranges <- boat_data %>%
    group_by(event_id) %>%
    summarize(start_time = min(time), end_time = max(time))
  
  # Initialize the event ID column for the full dataset
  gps_data$event_id <- NA
  
  # Loop through each event range and label the points in the full dataset
  for (i in 1:nrow(event_ranges)) {
    gps_data$event_id[gps_data$time >= event_ranges$start_time[i] & gps_data$time <= event_ranges$end_time[i]] <- event_ranges$event_id[i]
  }
  
  # Filter out boat events with fewer than 4 boat points
  valid_event_ids <- boat_data %>%
    group_by(event_id) %>%
    filter(n() >= 6) %>%
    pull(event_id) %>%
    unique()
  
  boat_data_filtered <- gps_data %>%
    filter(event_id %in% valid_event_ids)

  return(boat_data_filtered)
}

```

# identify GPS track segments with a back and forth movement pattern in them
```{r}
extract_boat_events <- function(boat_data_filtered) {
  # Initialize a column to store if the event is a "boat event" based on distance
  boat_data_filtered$is_boat_event <- FALSE
  
  # Function to calculate Euclidean distance between two UTM points
  euclidean_distance <- function(x1, y1, x2, y2) {
    sqrt((x2 - x1)^2 + (y2 - y1)^2)
  }
  
  unique_events <- unique(boat_data_filtered$event_id)
  
  # Loop through each event
  for (id in unique_events) {
    
    # Extract the event data for this event_id, excluding rows with NA coordinates
    event_data <- boat_data_filtered %>%
      filter(event_id == id & !is.na(x) & !is.na(y))
    
    # Skip if event_data is empty
    if (nrow(event_data) == 0) {
      next
    }
    
    # Calculate the actual traveled distance (along the whole track)
    total_traveled_distance <- 0
    for (i in 2:nrow(event_data)) {
      total_traveled_distance <- total_traveled_distance + euclidean_distance(
        event_data$x[i-1], event_data$y[i-1], event_data$x[i], event_data$y[i]
      )
    }
    
    # Calculate the convex hull and the absolute distance (size of the track in space)
    coords <- as.matrix(event_data[, c("x", "y")])
    if (nrow(coords) > 2) {  # Ensure there are enough points to form a hull
      hull <- chull(coords)  # Convex hull indices
      hull_coords <- coords[hull, ]  # Convex hull coordinates
      
      # Calculate the perimeter of the convex hull
      hull_perimeter <- 0
      for (i in 2:nrow(hull_coords)) {
        hull_perimeter <- hull_perimeter + euclidean_distance(
          hull_coords[i-1, 1], hull_coords[i-1, 2], hull_coords[i, 1], hull_coords[i, 2]
        )
      }
      
      # Close the convex hull loop (distance between the last and first points)
      hull_perimeter <- hull_perimeter + euclidean_distance(
        hull_coords[nrow(hull_coords), 1], hull_coords[nrow(hull_coords), 2], hull_coords[1, 1], hull_coords[1, 2]
      )
      
      # Compare actual traveled distance to the convex hull size
      if (total_traveled_distance > 1.3 * hull_perimeter) {
        boat_data_filtered$is_boat_event[boat_data_filtered$event_id == id] <- TRUE
      }
    }
  }

  selected_events <- boat_data_filtered %>% filter(is_boat_event == TRUE)
  return(selected_events)
}
```

# run functions on GPS data to extract boat events
```{r}
# create link to data files
file_list <- list.files(path = "HMM_processed_data", pattern = ".csv", full.names = TRUE)

all_boat_events <- data.frame() # store selected GPS track segments
failed_files <- c() # store files that can't be processed

# function call
for (i in 1:length(file_list)){
  
  tryCatch({
  
    # read in data
    gps_data <- read.csv(file_list[i])
    gps_data$time <- as.POSIXct(gps_data$time, tz = "UTC", format = "%Y-%m-%d %H:%M:%S")
    
    # Ensure data format is similar across files and filter NA values
    gps_data[,3:23] <- as.character(gps_data[,3:23])
    event_data <- filter_states(gps_data)
    
    # extract GPS track segments
    selected_events <- extract_boat_events(event_data)
    
    if (nrow(selected_events) > 0){
      all_boat_events <- bind_rows(all_boat_events, selected_events)
      
      print(paste("Done with", file_list[i]))
    } else {
      print(paste("No events for", file_list[i]))
    }
  
  }, error = function(e) {
    # Print error message and continue with the next item
    print(paste("Error processing", file_list[i], ":", e$message))
  })

}

# check number of unique events
count_combinations <- n_distinct(all_boat_events$device, all_boat_events$event_id)

# save selected data
write.csv(all_boat_events, "all_boat_events.csv", row.names = FALSE)
```

# check detection recall
```{r}
# Function to check if two time intervals overlap
time_overlap <- function(start1, end1, start2, end2) {
  return(start1 <= end2 & end1 >= start2)
}

# Function to check for overlaps between boat events and event summary, and log NA cases
check_event_overlap <- function(boat_events, event_summary_df) {
  
  # Add new columns to store the result (TRUE if overlap found, FALSE otherwise)
  boat_events$overlap <- FALSE
  event_summary_df$overlap <- FALSE
  
  # Add a column to store the Event nr. from boat_events for which the event overlaps
  event_summary_df$boat_event_nr <- NA
  
  # Create data frames to store cases where NA was found
  na_boat_events <- data.frame()
  na_matching_events <- data.frame()
  
  # Loop through each event in boat_events
  for (i in 1:nrow(boat_events)) {
    
    # Extract Device ID, Start time, End time, and Event nr. from the boat event
    device_id_boat <- boat_events$`Device ID`[i]
    start_time_boat <- as.POSIXct(boat_events$`Start time`[i], format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
    end_time_boat <- as.POSIXct(boat_events$`End time`[i], format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
    event_nr_boat <- boat_events$`Event nr.`[i]  # Extract the Event nr.
    
    # Check if boat event times are NA and log them
    if (is.na(start_time_boat) | is.na(end_time_boat)) {
      na_boat_events <- rbind(na_boat_events, boat_events[i, ])
      next  # Skip this boat event if start or end time is NA
    }
    
    # Filter event_summary_df for the same Device ID
    matching_events <- event_summary_df[event_summary_df$device == device_id_boat, ]
    
    # Check if matching_events is empty
    if (nrow(matching_events) == 0) {
      next  # Skip to the next boat event if no matching events found
    }
    
    # Check if there is any overlap with any event in the filtered event_summary_df
    for (j in 1:nrow(matching_events)) {
      
      start_time_summary <- as.POSIXct(matching_events$start_time[j], format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
      end_time_summary <- as.POSIXct(matching_events$end_time[j], format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
      
      # Check if summary times are NA and log them
      if (is.na(start_time_summary) | is.na(end_time_summary)) {
        na_matching_events <- rbind(na_matching_events, matching_events[j, ])
        next  # Skip this matching event if start or end time is NA
      }
      
      # Check if times overlap using the custom function
      if (time_overlap(start_time_boat, end_time_boat, start_time_summary, end_time_summary)) {
        # Mark overlap in boat_events
        boat_events$overlap[i] <- TRUE  
        
        # Mark overlap in event_summary_df
        match_index <- which(event_summary_df$device == device_id_boat & 
                             event_summary_df$start_time == matching_events$start_time[j] & 
                             event_summary_df$end_time == matching_events$end_time[j])
                             
        event_summary_df$overlap[match_index] <- TRUE
        event_summary_df$boat_event_nr[match_index] <- event_nr_boat  # Add the Event nr. to the matching event
        
        break  # If overlap found, stop checking further events for this boat event
      }
    }
  }
  
  # Return the result and NA cases
  return(list(boat_events_overlap = boat_events, event_summary_overlap = event_summary_df, 
              na_boat_events = na_boat_events, na_matching_events = na_matching_events))
}

# Example usage of the function
result <- check_event_overlap(boat_events_hand_picked, event_summary_df)

# Get the boat events with overlap
boat_events_with_overlap <- result$boat_events_overlap
event_summary_with_overlap <- result$event_summary_overlap

table(boat_events_with_overlap$overlap)
```

# visually check detection and classification precision
```{r}
plot_selected_events <- function(gps_track, selected_events) {
  
  # Extract device ID from GPS track data
  device_ID <- as.numeric(gps_track$device[1])
  
  # Filter the selected events to only include events for the current device
  df_events <- selected_events[!is.na(selected_events$device) & selected_events$device == device_ID, ]
  
  # Convert the time column in GPS data to POSIXct format for later filtering
  gps_track$time <- as.POSIXct(gps_track$time, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
  
  # Check if there are any relevant events for the current device
  if (nrow(df_events) == 0) {
    print(paste("No selected events for device", device_ID))
  
  } else {
    
    # Define the state names for different behaviors or activities
    state_names <- c("Stationary", "Float", "ARS", "Exploratory Flight", "Transit", "Trawler", "Shrimp boat")
    
    # Define a named vector for colors corresponding to the states
    state_colors <- c("Stationary" = "red", "Float" = "blue", "ARS" = "green",
                      "Exploratory Flight" = "purple", "Transit" = "orange",
                      "Trawler" = "brown", "Shrimp boat" = "pink")
    
    # Loop through each selected event for the current device
    for (i in 1:nrow(df_events)) {
      
      # Extract start and end times for the current event
      start_time <- as.POSIXct(df_events$start_time[i], tz = "UTC")
      end_time <- as.POSIXct(df_events$end_time[i], tz = "UTC")
      
      # Filter GPS data to include only points within the event time range
      filtered_data <- gps_track %>% filter(time >= start_time & time <= end_time)
      
      # Check if filtered_data has any points to plot
      if (nrow(filtered_data) > 0) {
        # Create the plot for the current event
        p <- ggplot(filtered_data, aes(x = x, y = y, color = States)) +
          geom_point() +
          scale_color_manual(values = state_colors) + # Use custom colors for states
          labs(title = paste("Track/Point Plot", device_ID, "- Event", df_events$event_id[i], "duration", as.numeric(df_events$event_duration[i]) / 60), # Add event index to the title
               x = "X Coordinate",
               y = "Y Coordinate",
               color = "States") +
          geom_segment(aes(xend = lead(x), yend = lead(y)), arrow = arrow(length = unit(0.2, "cm"))) + 
          theme_minimal()
        
        # Print plot
        print(p)
      } else {
        print(paste("No GPS points in the range for event", i, "for device", device_ID))
      }
    }
  }
}


# Randomly select 50 numbers from 1 to 463
set.seed(1)  # Setting seed for reproducibility
random_numbers <- sample(1:463, 50)

# Subset event_summary_df based on the selected random indices
precision_plot_data <- event_summary_df[random_numbers, ]

# Load GPS data files and call plot_selected_events for each file
file_list <- list.files(path = "HMM_processed_data", pattern = ".csv", full.names = TRUE)

for (i in 1:length(file_list)){
  
  gps_data <- read.csv(file_list[i])
  
  # Pass precision_plot_data to plot_selected_events instead of the full boat_events_hand_picked
  plot_selected_events(gps_data, precision_plot_data)

}

```