---
title: "HMM model iterate"
output: html_notebook
---

This file contains code to run a Hidden Markov Model on GPS tracks, which uses step lengths and turning angles to determine different behavioral states along the track. The code performs 15 iterations of the HMM with different starting values each time. It then selects and stores the output of the best fitted model.

The input CSV file should have the following columns:
1. ID (string or integer): Unique identifier for each device or bird.
2. track_ID (string or integer): Unique identifier for each track segment.
3. time (datetime): Timestamp of each data point in the format "YYYY-MM-DD HH:MM:SS".
4. x (numeric): UTM x-coordinate of the data point.
5. y (numeric): UTM y-coordinate of the data point.

This code was written by Dr. E.N. (Eldar) Rakhimberdiev and adapted by Eva de Jonghe. The script is based on this paper: https://doi.org/10.1111/2041-210X.13745 by Cullen et al., 2021
but with the extended code for the momentuHMM package.https://doi.org/10.1111/2041-210X.12995 by McClintock and Théo Michelot, 2017.

```{r}
# Load required libraries
library(momentuHMM)
library(bayesmove)
library(tidyverse)
library(terra)
```

# prepare data for HMM model
```{r}
# read in data
gullsubs <- read.csv("C:/Users/edejonghe/Documents/Thesis/7062_data_010122_311222_processed.csv")

# change the "ID" column so the HMM takes separate tracks into account
gullsubs <- gullsubs %>% rename("device" = "ID", "ID" = "track_ID")

# if HMM gives the error that there are not enough input values --> remove duplicates
#gullsubs <- gullsubs[!duplicated(gullsubs[2]), ] 

# for subsetting large datasets
#gullsubs <- gull5999 |> subset(date_time > as.POSIXct('2021-04-09 00:00:00', tz="UTC") & date_time < as.POSIXct('2021-07-09 00:00:00', tz="UTC")) # this one worked well for me...

# transform data into lists for the HMM model
d.list<- df_to_list(gullsubs, ind = "device")
list1<- map(d.list, ~prepData(., type = "UTM", coordNames = c("x","y")))
```

# Fit HMM model
```{r}
#############################
# Here the HMM model will be run for 15 iterations with different initial values.

# Empty list for order selection
k.models<- list()

## K = 5
allm<- list()
niter<- 15
stateNames <- c("Stationary", "Float","ARS", "Exploratory Flight", "Transit", "Trawler", "Shrimp boat")
whichzero <- which(list1[[1]]$step == 0)
propzero <- length(whichzero)/nrow(list1[[1]])
zeromass0 <- c(propzero, propzero, 0, 0)        #for zero distances by state
    
for (i in 1:niter) {
  print(paste("K=7"))
  print(paste("Iteration", i))
      
  # Mean Step length for different states
  stepMean0 <- runif(7, min = c(0.01, 50,  10,  500, 2000, 800, 400),
                        max = c(5, 300, 500, 2000, 5000, 1000, 500))

  # Step length standard deviation
  stepSD0 <- runif(7, min = c(0.1, 10, 50, 500, 500, 11, 5),
                      max = c(1, 100, 250, 1500, 1500, 101, 50))


  # Mean turning angle for different states
  angleMean0 <- c(pi, 0, pi, 0, 0, pi, pi)
      
  # Turning angle concentration
  angleCon0 <- runif(7, min = c(0.01, 0.8, 0, 0.1, 0.5, 0, 0),
                        max = c(0.5, 0.99, 0.5, 0.5, 0.99, 0.6, 0.6))
      
  # Fit HMM model
  if(propzero > 0) {  #don't include zero mass if no 0s present
        
    stepPar0 <- c(stepMean0, stepSD0, zeromass0)
      
    } else {
        
      stepPar0 <- c(stepMean0, stepSD0)
      
      }
      
  anglePar0 <- c(angleMean0, angleCon0)
	  
  start.time<- Sys.time() # measure elapsed time per iteration

  Trial <- fitHMM(data = list1[[1]], nbStates = 7,
                  Par0 = list(step = stepPar0, angle = anglePar0),
                  dist = list(step = "gamma", angle = "wrpcauchy"),
                  formula = ~ 1, stationary=TRUE, #stationary for a slightly better fit
                  estAngleMean = list(angle=TRUE),
                  stateNames = stateNames,
                  optMethod = "Nelder-Mead") 
	 
  end.time<- Sys.time()
  elapsed.time<- difftime(end.time, start.time, units = "min")
  cat  ('time elapsed:\n')
	print(elapsed.time)
     
	if(!inherits(Trial, "try-error")) {
      
	  allm[[i]]<-Trial 
    print(Trial) }
    
	}
   
# save model output
save(allm, file='first_run_5_states_15_models_3_monts.RData')
# allm_old<-allm

# Extract likelihoods of fitted models
allnllk_list<-lapply(allm, function(m) m$mod$minimum)
allnllk_list[unlist(lapply(allnllk_list, is.null))]<-NA
allnllk <- unlist(allnllk_list)

# Index of best fitting model (smallest negative log-likelihood)
whichbest <- which.min(allnllk)
    
# Best fitting model
best_model<- allm[[whichbest]]
# save best model data
save(best_model, file = 'best_model_080724.RData')
saveRDS(best_model, "best_model_080724.rds") #saves the model in device

```

# save outcome
```{r}
# extract the states found by the HMM
States <- viterbi(best_model)
gullsubs$States <- States # add them to dataset

write.csv(gullsubs,"C:/Users/edejonghe/Documents/Thesis/7062_data_010122_311222_processed.csv", row.names = FALSE) # save dataset

```

# plot outcome    
```{r}
# model output and diagnostics
plot(best_model)
 
plot(Trial)

plotPR(best_model)

# plot (part of) tracks with predictions
par(mfrow=c(2,1))
Colors=c('red','brown', 'pink', 'orange', 'blue', 'green', 'darkgreen')
plot(gullsubs$y~gullsubs$time, type='l')
points(gullsubs$y~gullsubs$time, pch='+', col=Colors[States])

plot(gullsubs$x~gullsubs$date_time, type='l')
points(gullsubs$x~gullsubs$date_time, pch='+', col=Colors[States])

plot(gullsubs$dist~gullsubs$date_time, type='l')
points(gullsubs$dist~gullsubs$date_time, pch='+', col=Colors[States])

#
cur_set<-1:50
cur_set<-301:600
Colors=c('red','brown', 'pink', 'orange', 'blue', 'green', 'darkgreen')
plot(gullsubs$y[cur_set]~gullsubs$time[cur_set], type='l')
points(gullsubs$y[cur_set]~gullsubs$time[cur_set], pch='+', col=Colors[States[cur_set]])

plot(gullsubs$x[cur_set]~gullsubs$date_time[cur_set], type='l')
points(gullsubs$x[cur_set]~gullsubs$date_time[cur_set], pch='+', col=Colors[States[cur_set]])

plot(gullsubs$y[cur_set]~gullsubs$x[cur_set], type='l')
points(gullsubs$y[cur_set]~gullsubs$x[cur_set], pch='+', col=Colors[States[cur_set]])

legend(x = "topleft",          # Position
       legend = stateNames,  # Legend texts
       lty = c(1),           # Line types
       col = Colors,           # Line colors
       lwd = 1)                 # Line width
```
